RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R

### 2014-04-28 inst/doc/pubs/manuscript.Rmd
Update manuscript.Rmd

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NBC
    address: | 
      NBC Naturalis, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange format, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as NEXUS.

      2. In this paper we present the RNeXML package which provides
      a bridge between this rich and reliable data format and the
      extensive library of phylogenetic tools already available in R,
      with a particular emphasis on comparative methods.

@@

* Why NeXML -- because NEXUS standard isn't a standard.  Developers extend
it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
For example, NEXUS cannot represent horizontal gene transfer, ambiguous
use of symbols, cannot be extended reliably, and there is no way to
validate that it is proper NEXUS.  NeXML addresses all of these problems
[Vos _et al._ 2012] with a state-of-the-art data management format.

@@

A nexml file can be thought of as having three components: (1) phylogeny
data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
provide a seemless way to convert these to and from their corresponding
R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.

>>>>

---
layout: preprint
#layout: review, 11pt
title: "RNeXML: Parsing and Serializing the Next Generation of Phyloinformatic Data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig@gmail.com
    footnote: Corresponding author
  - name: Hilmar Lapp
    affiliation: NESCent
  - name: Scott Chamberlain
    affiliation: SimonFraser
  - name: Rutger Vos
    affiliation: NBC
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center, Durham, NC, USA
  - code: SimonFraser
    address: |
      Department of Biology, Simon Fraser University, Burnaby, Canada
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML promises to be the next generation of phylogenetic
      informatics data and metadata exchange formats, offering substantial
      improvements in reliability, extensibility, and richness over
      earlier standards such as NEXUS.

      2. In this paper we present the RNeXML package, which provides
      a bridge between this rich and reliable data format and the
      extensive library of phylogenetic tools already available in R,
      with a particular emphasis on comparative methods.

@@

* Why NeXML -- because the NEXUS standard isn't a standard.  Developers extend
it arbitrarily to meet their needs, yielding mutually incompatible dialects. 
For example, NEXUS cannot represent horizontal gene transfer, ambiguous
use of symbols, cannot be extended reliably, and there is no way to
validate that it is proper NEXUS.  NeXML addresses all of these problems
[Vos _et al._ 2012] with a state-of-the-art data management format.

@@

A NeXML file can be thought of as having three components: (1) phylogeny
data (2) character-trait data, and (3) metadata. `RNeXML` seeks to
provide a seemless way to convert these to and from their corresponding
R object representations, such as `ape::phylo` phylogenies, character
matrices (`data.frames`) and metadata lists (or for advanced users,
`XMLInternalDocument` and `rrdf` triplestores). Metadata can provide
information about the study in which the phylogeny was generated or
applied (such as authors, citations to publications, timestamps, and even
descriptions of methods), but can also be used to annotate individual
data components.  We illustrate how metadata annotations can be added to
the operational taxonomic units of the phylogeny and can be used to extend
the data representation to new uses such as stochastic character maps.
